---
categories:
- ""
- ""
date: "2020-10-30"
description: Us Yield Curve Reversion
draft: false
keywords: ""
slug: usyield
title: Us Yield Curve Reversion
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```


First, we will use the `tidyquant` package to download monthly rates for different durations. 

```{r get_rates, warning=FALSE}
# Get a list of FRED codes for US rates and US yield curve; choose monthly frequency
# to see, eg., the 3-month T-bill https://fred.stlouisfed.org/series/TB3MS
tickers <- c('TB3MS', # 3-month Treasury bill (or T-bill)
             'TB6MS', # 6-month
             'GS1',   # 1-year
             'GS2',   # 2-year, etc....
             'GS3',
             'GS5',
             'GS7',
             'GS10',
             'GS20',
             'GS30')  #.... all the way to the 30-year rate

# Turn  FRED codes to human readable variables
myvars <- c('3-Month Treasury Bill',
            '6-Month Treasury Bill',
            '1-Year Treasury Rate',
            '2-Year Treasury Rate',
            '3-Year Treasury Rate',
            '5-Year Treasury Rate',
            '7-Year Treasury Rate',
            '10-Year Treasury Rate',
            '20-Year Treasury Rate',
            '30-Year Treasury Rate')

maturity <- c('3m', '6m', '1y', '2y','3y','5y','7y','10y','20y','30y')

# by default R will sort these maturities alphabetically; but since we want
# to keep them in that exact order, we recast maturity as a factor 
# or categorical variable, with the levels defined as we want
maturity <- factor(maturity, levels = maturity)

# Create a lookup dataset
mylookup<-data.frame(symbol=tickers,var=myvars, maturity=maturity)
# Take a look:
mylookup %>% 
  knitr::kable()

df <- tickers %>% tidyquant::tq_get(get="economic.data", 
                   from="1960-01-01")   # start from January 1960

glimpse(df)
```

Our dataframe `df` has three columns (variables):

- `symbol`: the FRED database ticker symbol
- `date`: already a date object
- `price`: the actual yield on that date

The first thing would be to join this dataframe `df` with the dataframe `mylookup` so we have a more readable version of maturities, durations, etc.


```{r join_data, warning=FALSE}

yield_curve <-left_join(df,mylookup,by="symbol") 
```

## Plotting the yield curve

This may seem long but it should be easy to produce the following three plots

### Yields on US rates by duration since 1960
```{r}
yield_curve$var<-factor(yield_curve$var, levels = myvars)

```

```{r}
yield_curve %>% 
  ggplot(aes(x=date, y=price,colour=maturity))+
  geom_line()+
  facet_wrap(~var,nrow=5)+
  theme_bw()+
  theme(legend.position = 'none')+
  labs(title='Yields on U.S. Treasury rates since 1960',
       x=element_blank(),
       y='%',
       caption='Source: St Louis Federal Reserve Economic Database(FRED)')

```

### Monthly yields on US rates by duration since 1999 on a year-by-year basis

```{r}
yield_curve %>% 
  mutate(year=year(date),month=month(date)) %>% 
  filter(year%in%c(1999:2020)) %>% 
  ggplot(aes(x=maturity, y=price,group=month, colour=year))+
  geom_line()+
  facet_wrap(~year,nrow=5)+
  theme_bw()+
  theme(legend.position = 'none')+
  labs(title='US Yield Curve',
       x='Maturity',
       y='Yield(%)',
       caption='Source: St Louis Federal Reserve Economic Database(FRED)')
```

### 3-month and 10-year yields since 1999
```{r}
yield_curve %>% 
  mutate(year=year(date)) %>% 
  filter(year%in% c(1999:2020)& maturity%in% c('3m','10y')) %>% 
  ggplot(aes(x=date, y=price, colour=var))+
  geom_line()+
  theme_bw()+
  labs(title='US Yield Curve',
       x='Maturity',
       y='Yield(%)',
       caption='Source: St Louis Federal Reserve Economic Database(FRED)')
```




According to [Wikipedia's list of recession in the United States](https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States), since 1999 there have been two recession in the US: between Mar 2001–Nov 2001 and between Dec 2007–June 2009. Does the yield curve seem to flatten before these recessions? Can a yield curve flattening really mean a recession is coming in the US? Since 1999, when did short-term (3 months) yield more than longer term (10 years) debt?

```{r}

#Yes, the yield curve flatten several months earlier than rec，ssions, which is a very useful indicator for recession.
#from Aug 2000 to Dec 2000 and from Aug 2006 to Apr 2007.
```



Besides calculating the spread (10year - 3months), there are a few things we need to do to produce our final plot

1. Setup data for US recessions
1. Superimpose recessions as the grey areas in our plot
1. Plot the spread between 30 years and 3 months as a blue/red ribbon, based on whether the spread is positive (blue) or negative(red)


- For the first, the code below creates a dataframe with all US recessions since 1946

```{r setup_US-recessions, warning=FALSE}

# get US recession dates after 1946 from Wikipedia 
# https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States

recessions <- tibble(
  from = c("1948-11-01", "1953-07-01", "1957-08-01", "1960-04-01", "1969-12-01", "1973-11-01", "1980-01-01","1981-07-01", "1990-07-01", "2001-03-01", "2007-12-01"),  
  to = c("1949-10-01", "1954-05-01", "1958-04-01", "1961-02-01", "1970-11-01", "1975-03-01", "1980-07-01", "1982-11-01", "1991-03-01", "2001-11-01", "2009-06-01") 
  )  %>% 
  mutate(From = ymd(from), 
         To=ymd(to),
         duration_days = To-From)

recessions
```

- To add the grey shaded areas corresponding to recessions, we use `geom_rect()`

```{r}
library(scales)
```

```{r}
yield_3m=filter(yield_curve,maturity=='3m')
yield_10y=filter(yield_curve,maturity=='10y')

yield_diff<-yield_3m %>% 
  mutate(price10=yield_10y$price,
         diff=price10-price,
         judge=diff>0,
         posi=ifelse(judge,diff,0),
         nega=ifelse(judge,0,diff)) 

yield_diff %>% 
  ggplot(aes(x=date))+
  geom_area(aes(y=posi,fill='red'),outline.type='lower')+
  geom_area(aes(y=nega,fill='blue'),outline.type='lower')+
  geom_rect(data=filter(recessions,year(From)>1960),inherit.aes=F, aes(xmin=From, xmax=To, ymin=-Inf, ymax=+Inf), fill='grey', alpha=0.5)+
  geom_line(aes(y=diff))+
  geom_rug(sides="b",aes(colour=judge)) +
  theme_bw()+
  theme(legend.position = 'none')+
  labs(title='Yield Curve Inversion: 10-year minus 3-month U.S. Treasury rates',
       subtitle='Difference in % points, monthly averages.
       shaded areas correspond to recessions',
       x=element_blank(),
       y='Difference(10 year-3 month) yield in %',
       caption='Source: FRED, Federal Reserve Bank of St Louis')
```
