---
categories:
- ""
- ""
date: "2020-10-30"
description: Excess Rentals in London Bike Sharing
draft: false
keywords: ""
slug: bike
title: Excess Rentals in London Bike Sharing
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(scales)
```


Since bike sharing comes to popular, it has become a  reflection of people's activities and can show the impact of those great event like covid-19 this year. In this analysis, we will download bike sharing data from UK gov, and observe the different patterns of monthly bike-hire number between 2014-2016. We use the average number of each month as the expected rental number, and calculate the difference from expected rental number and read rental number. And then we will conduct the same method to analyze weekly data.


```{r, get_tfl_data, cache=TRUE}

url <- "https://data.london.gov.uk/download/number-bicycle-hires/ac29363e-e0cb-47cc-a97a-e216d900a6b0/tfl-daily-cycle-hires.xlsx"

# Download TFL data to temporary file
httr::GET(url, write_disk(bike.temp <- tempfile(fileext = ".xlsx")))

# Use read_excel to read it as dataframe
bike0 <- read_excel(bike.temp,
                   sheet = "Data",
                   range = cell_cols("A:B"))
Sys.setlocale("LC_TIME", "English")
# change dates to get year, month, and week
bike <- bike0 %>% 
  clean_names() %>% 
  rename (bikes_hired = number_of_bicycle_hires) %>% 
  mutate (year = year(day),
          month = lubridate::month(day,label=TRUE),
          week = isoweek(day))
```

```{r}
Sys.setlocale("LC_TIME", "English")
```

## Data Processing

```{r}
bike2<-bike %>% 
  group_by(year,month) %>% 
  summarise(monthly_num=mean(bikes_hired)) %>% 
  ungroup()

bike3<-bike2%>% 
  filter(year>2014&year<2020) %>% 
  group_by(month) %>% 
  summarise(monthly_mean=mean(monthly_num)) %>% 
  ungroup()

bike4<-left_join(bike2,bike3,"month")%>% 
  mutate(posi=ifelse(monthly_num>monthly_mean,monthly_num,monthly_mean),
         nega=ifelse(monthly_num>monthly_mean,monthly_mean,monthly_num))%>% 
  filter(year>2014) 

```


```{r}
bike_w<-bike %>% 
  group_by(year,week) %>% 
  summarise(weekly_num=mean(bikes_hired))%>% 
  ungroup()

bike_w2<-bike_w%>% 
  filter(year>2014&year<2020) %>% 
  group_by(week) %>% 
  summarise(weekly_mean=mean(weekly_num)) %>% 
  ungroup()

bike_w3<-left_join(bike_w,bike_w2,"week")%>% 
  mutate(diff=(weekly_num-weekly_mean)/weekly_mean,
        judge=weekly_num>weekly_mean,
        posi=ifelse(judge,diff,0),
         nega=ifelse(judge,0,diff))%>% 
  filter(year>2014) 


```


## Monthly changes in Efl bike retals

```{r}
bike4 %>% 
  ggplot(aes(x=month))+
  geom_ribbon(aes(ymin=nega, ymax=monthly_num,fill='red',group=0))+
  geom_ribbon(aes(ymin=monthly_mean, ymax=nega,fill='bluw',group=0))+
  geom_line(aes(y=monthly_num),group=1)+
  geom_line(aes(y=monthly_mean),colour='#0101DF',size=1,group=1)+
  facet_wrap(~year)+
  theme_minimal()+
  labs(title='Monthly changes in TfL bike rentals',
       subtitle='Change from monthly average shown in blue and calculated between 2015-2019' ,
       y = "Bike Rentals",
       x=element_blank(),
       caption='Soource: TfL, London Data Store')+
  theme(legend.position ='none')

```

## Weekly changes in Efl bike retals

```{r}
ggplot(bike_w3,aes(x=week))+
  geom_area(aes(y=posi,fill='green'),outline.type='lower')+
  geom_area(aes(y=nega,fill='blue'),outline.type='lower')+
  geom_rect(aes(xmin=13, xmax=26, ymin=-Inf, ymax=Inf),fill='#545454',alpha = .005)+
  geom_rect(aes(xmin=39, xmax=52, ymin=-Inf, ymax=Inf),fill='#545454',alpha = .005)+
  geom_line(aes(y=diff))+
  facet_wrap(~year)+
  theme_minimal()+
  geom_rug(sides="b",aes(colour=judge)) +
  scale_y_continuous(labels = scales :: percent)+
  labs(title='Weekly changes in TfL bike rentals',
       subtitle='%changes from weekly averages calculated between 2015-2019',
       y=element_blank(),
       caption='Source: TfL, London Data Store')+
  scale_y_continuous(labels = percent,breaks=seq(-0.6,0.6,0.3))+
  scale_x_continuous(breaks=seq(0, 52, 13))+
  theme(legend.position="none")

```
